A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-07-15 paper.tex
thing
more free energy and cold/hot thermodynamic limits

\item
coding, transfer, and consolidation of
information
This emerging picture of hippocampal dy- namics suggests that neurons at the opposite ends of the distributions may convey differ- ent but complementary types of information. The ever-active minority of place cells may be responsible for generalizing across envi- ronments and affords the brain the capacity to regard no situation as completely un- known because every alley, mountain, river, or room has elements of previously experi- enced similar situations. In many situations,
this minority provides the “best guess” of the hippocampus and offers “good enough” solutions to get by. On the other hand, the majority of less active neurons constitute a large reservoir that can be mobilized (7) to precisely distinguish one situation from an- other and incorporate novel ones as distinct.

@@

\paragraph{How many controls do we need at best ?}
In the thermodynamic limit ($n \rightarrow \infty$), statistical physics \cite{Liu2011} gives the extremely good estimate
\begin{equation}
  n_0(\langle k\rangle, \gamma) \approx e^{-\frac{1}{2}\left(\frac{\gamma-2}{\gamma - 1}\right)\langle k \rangle},
\end{equation}
where $\gamma \ge 1$ is the \textit{scale-free} parameter for the node-degree distribution of the network described by $\A$, and $\langle k \rangle$ is the mean degree. Letting $\gamma \rightarrow \infty$, one recovers the Erdos-Renyi value
\begin{equation}
  n_0(\langle k\rangle) = e^{-\frac{1}{2}\langle k\rangle},
\end{equation}
which agrees perfectly with its known analytic value. One identifies two radically different regimes:
\begin{itemize}
  \item the manageable regime $\gamma > \gamma_c := 2$, in which $n_0 < 1$, and so we only need fewer controls than total number of nodes, and
\item the unmanageable regime ``$\gamma \le \gamma_c$'', in which  $n_0 = 1$, where each node must be controlled explicitly.
\end{itemize}

@@

\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}

>>>>

\item
coding, transfer, and consolidation of
information
This emerging picture of hippocampal dy- namics suggests that neurons at the opposite ends of the distributions may convey differ- ent but complementary types of information. The ever-active minority of place cells may be responsible for generalizing across envi- ronments and affords the brain the capacity to regard no situation as completely un- known because every alley, mountain, river, or room has elements of previously experi- enced similar situations. In many situations,
this minority provides the “best guess” of the hippocampus and offers “good enough” solutions to get by. On the other hand, the majority of less active neurons constitute a large reservoir that can be mobilized (7) to precisely distinguish one situation from an- other and incorporate novel ones as distinct.
\end{itemize}
\end{itemize}

@@

\paragraph{How many controls do we need at best ?}
In the thermodynamic limit ($n \rightarrow \infty$), statistical physics \cite{Liu2011} gives the extremely good estimate
\begin{equation}
  n_0(\langle k\rangle, \gamma) \approx \exp\left(-\frac{1}{2}\left(\frac{\gamma-2}{\gamma - 1}\right)\langle k \rangle\right),
\end{equation}
where $\gamma \ge 1$ is the \textit{scale-free} parameter for the node-degree distribution of the network described by $\A$, and $\langle k \rangle$ is the mean degree. Letting $\gamma \rightarrow \infty$, one recovers the Erdos-Renyi value
\begin{equation}
  n_0(\langle k\rangle) = \exp\left(-\frac{1}{2}\langle k\rangle\right),
\end{equation}
which agrees perfectly with its known analytic value. One identifies two radically different regimes:
\begin{itemize}
  \item the manageable regime $\gamma > \gamma_c := 2$, in which $n_0 < 1$, and so we only need fewer controls than total number of nodes, and
\item the unmanageable regime ``$\gamma \le \gamma_c$'', in which  $n_0 = 1$, where each node must be controlled explicitly.
\end{itemize}

@@

\subsection{A thermodynamic model for bounded rationality, aka robust optimality}
\begin{itemize}
  \item Recall that an agent is said to have \textit{bounded rationality} if they must take into account the cost of finding solutions to problems, and not just the utility of the final state.
  \item For example, consider an agent that must operate under a limited lifetime and/or computation cost.
  \item This is in contrast to agents with \textit{unbounded rationality} considered in classical game theory.
    \end{itemize}
The material presented is a revisit of \textcolor{blue}{[Braun 2011]}.
\paragraph{Utility functions and conjugate pairs.}
  Let $(\Omega, \mathcal F, \P)$ be a probablity space. A function $\U: \mathcal F \rightarrow \mathbb R$ is said to be a \textit{utility function} for this space if the conditional utility $\U(A|B) := \U(A \cap B) - \U(B)$ has the following propertites:
  \begin{itemize}
  \item additivity: $\U(A_1 \cap A_2 | B) = \U(A_1|B) + \U(A_2|B)$, for all events $A_1, A_2, B \in \mathcal F$.    
  \item statistic: there exists a function $f_{\U} :\mathbb R_+ \rightarrow \mathbb R$ such that $\U(A|B) = f(\P(A|B))$, for all events
    $A,B \in \mathcal F$.
  \item monotonicity: $f_{\U}$ is strictly increasing.
    \end{itemize}

\begin{theorem}
  The only functions $f: \mathbb R_+ \rightarrow \mathbb R$ which is such that $\U(A|B) \equiv f(\P(A|B))$ any probability space $(\Omega, \mathcal F, \P)$ and utility function $\U$ thereupon are of the form
  \begin{equation}
    f = \alpha \log(.),
    \label{eq:boltzmann}
  \end{equation}
  where $\alpha > 0$.
\end{theorem}

\textbf{XXX: equation \eqref{eq:boltzmann} above looks like Boltzmann's formula (on his gravestone...)!}

Such $\U$ and $\P$ are said to form a \textit{conjugate pair} at temperature $\alpha$.

\paragraph{Example.}
Given a utility function $\U$ on a probability space $(\Omega, \mathcal F, *)$, the \textit{Gibbs measure} at temperature $\alpha > 0$ and energy levels $(-\U(\omega))_{\omega \in \Omega}$ is defined to be the probability measure (on thesame measurable space)
  \begin{equation}
    \P(A) = \sum_{\omega \in A}\frac{1}{Z(\alpha)}\exp\left(\frac{1}{\alpha}\U(\omega)\right), \; \forall A \in \mathcal F.
  \end{equation}
$Z(\alpha)$ is a normalization constant such that $\P(\Omega) = 1$.  It's not hard to see that $\U$ and the $\P$ above form a conjugate pair.

  \paragraph{Free-utility functional.}
    Let $(\U, \P)$ be a conjugate pair at temperature $\alpha > 0$  on a measurable space $(\Omega, \mathcal F)$. Given another probability measure $\P'$ on the same space, define it's \textit{free utility} as
    \begin{equation}
      J(\P'|\U, \P) = \mathbb E_{\P'}(\U) + \alpha \mathbb H(\P'),
    \label{eq:free_u}
    \end{equation}
    where $\mathbb H(\P') := -\sum_{\omega \in \Omega}\P'(\omega)\log(\P'(\omega))$ is the entropy of $\P'$.

  It's not difficult to show that
  \begin{equation}
    J(\P'|\U,\P) \le J(\P|\U,\P) = \sum_{\omega \in \Omega}\U(\omega) =: \U(\Omega).
  \end{equation}
  In particular, if $\P$ is the Gibbs measure corresponding to $\U$, then the upper bound above reduces to the \textit{log-partition function}
  \begin{equation}
    J(\P'|\U,\P) \le \U(\Omega) = -\alpha \log\left(\sum_{\omega \in \Omega}\exp\left(\frac{1}{\alpha}\U(\omega)\right)\right).
    \end{equation}

  \paragraph{The free-energy / utility principle.}
  We are now in shape to introduce the notion of free-energy for model transitions, and a variational principle for optimizing it. Consider thus an initial system described by a conjugate pair $(\U_{\text{ini}}, \P_{\text{ini}})$ at temperature $\alpha > 0$. We want to transform this to a new model by adding constraints represented by the utility function $\delta \U$.  The resulting system has final utility $\U_{\text{fin}} = \U_{\text{ini}} + \delta\U$. The difference in free-utility is then
  \begin{equation}
    \delta J_{(\U_{\text{ini}}, \P_{\text{ini}}) \rightarrow \P_{\text{fin}}} := J_{\text{fin}} - J_{\text{ini}} = \underbrace{\mathbb E_{\P_{\text{fin}}}(\delta \U)}_{\textbf{expected gain in utility}} + \alpha \underbrace{D_{\text{KL}}(\P_{\text{ini}}\|\P_{\text{fin}})}_{\textbf{complexity of change}},
    \label{eq:free}
  \end{equation}
  where $D_{\text{KL}}(\P_{\text{ini}}\|\P_{\text{fin}})$ is the Kullback-Leibler divergence.
  In the above formula, we've extensively used the fact that $(\U_{\text{ini}}, \P_{\text{ini}})$ is a congugate pair and so $\U_{\text{ini}}(\omega) \equiv \alpha \log(\P_{\text{ini}}(\omega))$ by virtue of \eqref{eq:boltzmann}.
  The two terms in \eqref{eq:free} (expected gain in utility or negative workdone, and the complexity of the transition) can be viewed as dertiminants of bounded rational decision-making. They formalize
a trade-off between an expected utility $\delta \U$ (first term) and
the information cost of transforming Pi
into $\P_{\text{fin}}$ (second
term). In this interpretation $\P_{\text{ini}}$ represents an initial choice
probability or policy, which includes the special case of the
uniform distribution where the decision-maker has initially no
preferences between the different choices. The probability measure $\P_{\text{fin}}$ is the final choice probability that we are looking for since it
considers the utility constraint U∗
that we want to optimize. We can then formulate a variational principle for bounded rationality in the probabilities $\P_{\text{fin}}(\omega)$
\begin{equation}
\P^*_{\text{fin}} := \underset{\P_{\text{fin}}}{\text{argmax }} \delta J_{(\U_{\text{ini}}, \P_{\text{ini}}) \rightarrow \P_{\text{fin}}}
  \end{equation}

By differentiating the RHS of \eqref{eq:free} w.r.t $\P_{\text{fin}}$ and setting to zero, we obtain the closed-form solution

\begin{equation}
  \P^*_{\text{fin}}(\omega) \propto \P_{\text{ini}}(\omega)\exp\left(\frac{1}{\alpha}\U(\omega)\right).
\end{equation}

Two limit cases are worth considering.
\begin{itemize}
\item \textbf{Low-temperature regime $\alpha \approx 0$:} Here $\delta J_{(\U_{\text{ini}}, \P_{\text{ini}}) \rightarrow \P_{\text{fin}}} \approx \mathbb E_{\P'_{\text{fin}}}(\delta \U)$, and so it's optimal to take $\P^*_{\text{fin}}(\omega) \equiv \text{dirac}(\omega - \omega^*)$, where $\omega^* := \argmax_{\omega \in \Omega}\U(\omega)$. This corresponds to unbounded rational decision-making, in which the cost of transition / problem-solving is completely disregarded.
  \item \textbf{High-temperature $\alpha \rightarrow +\infty$:} In this limiting case, it's optimal to take $\P^*_{\text{fin}}(\omega) \equiv \P_{\text{ini}}(\omega)$, i.e the change is so costly that it's optimal to maintain the current choice probabilities.
  \end{itemize}
\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}
